Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-07-26 manuscript.tex
post red-penning

\begin{abstract}
Ion cyclotron emission (ICE) originating from confined populations of fast ions
  in toroidal fusion plasmas is an important non-invasive, passive diagnostic
  for current and next generation devices. The ability to model the ICE signals
  accurately is an essential step towards inferring the characteristics of
  confined energetic alpha-particles or fast NBI ions and background plasma. In
  this paper, the linear growth rates of the magnetoacoustic cyclotron
  instability, which is the leading explanation for ICE, are calculated in high
  resolution 2D $(k_\perp, k_\parallel)$ space for parameters corresponding to
  JET Pulse No. 26148. These calculations shed further light on the origin of
  doublet-splitting of the cyclotron harmonics as well as the apparent need to
  invoke nonlinear interactions with 1D3V particle-in-cell studies to account
  for lower harmonics with substantial amplitudes: the inclusion of signals of
  ICE from a well resolved range of propagation angles can also account for
  these effects.
\end{abstract}

@@

Observations of ion cyclotron emission (ICE) have been reported from magnetically confined plasmas first in TFTR\cite{Cauffman1995,Dendy1995} and JET\cite{Cottrell1988,Cottrell1993,Dendy1995}, followed by other tokamaks PDX\cite{Heidbrink1994}, JT-60-U\cite{Ichimura2008}, DIII-D\cite{Heidbrink2011}, ASDEX\cite{DInca2014}, ASDEX-Upgrade\cite{Ochoukov2018}, EAST\cite{Liu2020}, and KSTAR\cite{Chapman2017,Chapman2018}. Stellarators\cite{Saito2013} and FRCs\cite{Nicks2020} have also exhibited this phenomenon. Energetic ion populations that give rise to ICE have their origins in fusion reactions in DT plasmas\cite{Cottrell2000, Cauffman1995}, ICRH\cite{Cottrell2000}, and neutral beam injection\cite{Ochoukov2018}. ICE arising from fusion $\alpha$-particles is a proposed route to further the understanding of burning plasmas in ITER\cite{Dendy2015,McClements2015}.

The foremost explanation of the mechanism responsible for ICE is the
magnetoacoustic cyclotron instability (MCI)\cite{Belikov1976}, which arises from
the free energy present in energetic ions with inverted distributions in
velocity space. Theoretical understanding of ICE is based in analytical
solutions to the linearised Maxwell-Vlasov system of equations\cite{Stix} where
the energetic ion species was modelled as ring-beam for which $f(v_\parallel,
v_\perp) \propto \exp(-(v_\parallel-u_\parallel)^2/ v_{th}^2)\delta(v_\perp -
u_\bot)$ \cite{Dendy1994, McClements1996}. Here $u_\parallel$, $u_\perp$ and
$v_{th}$ are the beam velocity (the signed drift speed along the magnetic
field), ring speed (the speed perpendicular to the magnetic field) and thermal
velocity, respectively. Subsequent studies augmented this understanding via
nonlinear fully-kinetic\cite{Cook2013,Chapman2017} and
hybrid\cite{Carbajal2014,Reman2019} particle-in-cell (PIC) simulations in one
spatial dimension and three vector components for velocities and electromagnetic
field (i.e. 1D3V), and recently 2D3V hybrid-PIC simulations\cite{Carbajal2021}. These
computational approaches operate under similar simplifications as the analytical
treatments in that the plasma is initially spatially homogeneous and periodic
in the spatial domain.
Linearised fluid approaches that incorporate toroidal geometry and
inhomogeneous plasma and magnetic fields have shed light of the eigenmode
structure of compressional Alfv{\'e}n eigenmodes, which like ICE are also fast
Alfv{\'e}n modes driven by energetic ions, in spherical
tokamaks\cite{Sharapov2014, Gorelenkov2016} and conventional aspect ratio
tokamaks \cite{Kolesnichenko2000, Fulop2000, Heidbrink2006}.

The effectiveness of analytical theory to reproduce the observed cyclotron
harmonic peaks in the fluctuations of the magnetic field was first explained
somewhat by 1D PIC simulations. In these simulations the onset of nonlinearity
appeared to imprint the amplitudes of each of the harmonics into the signal as
they are at the end of the linear phase; each mode grows independently until 
growth is terminated for all modes when the fastest mode enters its nonlinear
phase meaning that both the amount of free energy available for instability has
decreased and that the unstable energy ion distribution function has been
perturbed into a more stable state. Analysis shows that only higher cyclotron
harmonics are destabilised\cite{McClements2015} to the MCI under these
conditions, however experimental
observations clearly indicate power at theoretically stable lower
harmonics\cite{Cottrell1988}. In a further step towards accurate recreation of
experimental observations, nonlinear phases of hybrid\cite{Carbajal2014} and
fully kinetic\cite{Cook2013,Chapman2017} PIC simulations went on to show that
wave-wave coupling was responsible for raising the amplitude of lower cyclotron
harmonics. Most recently, hybrid 2D simulations have begun to shed light on the
2D3V dynamics of ICE\cite{Carbajal2021}.

Finer detailed structure has been obvserved in ICE spectra, such as doublet
splitting of individual harmonics on JET\cite{Cottrell1993} and
DIII-D\cite{Thome2018} and even
multiplet splitting on TUMAN-3M\cite{Askinazi2018} on EAST\cite{Lui2020}.
This paper highlights one underlying kinetic mechinism that produces
doublet splitting due to Doppler shifted resonances. In general,
splitting could be produced by grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998} or potentially any mechanism that is capable
of adding a Doppler shift to the wave-particle resonance condition.

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. In this work, I present
the growth rates in 2D wavenumber space of the MCI under the conditions of
outer mid-plane of the JET PTE\cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
LinearMaxwellVlasov.jl\cite{LinearMaxwellVlasov.jl}, written in the Julia programming
language\cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run readily on desktop computers
in a matter of minutes, further enabling research into this topic. Note that
this study focusses on edge-ICE, named for the region of plasma from which the signal
arises, as opposed to core-ICE\cite{Ochoukov2018}. Recent observations made with
higher resolution equipment than those referred to on JET have
shown multiplet splitting in NBI-driven spectra; from the core on
ASDEX-Upgrade\cite{Ochoukov2019} and edge on EAST\cite{Lui2020}.

In section 2, the computational method is described for calculating the linear growth rates, which are presented for energetic ring-beam alpha-particles with non-zero perpendicular thermal spread in section 3. Results are summarised in section 4.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are found in
parallel and perpendicular wavenumber space, $(k_\parallel, k_\bot)$, for a
ring-beam distribution of energetic minority
alpha-particles and Maxwellian electrons and background thermal majority
deuteron ions.

The ring-beam distribution function, Eq. 1, is a separable
function of the beam in $v_\parallel$ given by Eq. 2 and the
ring in $v_\perp$ given by Eq. 3.
In the context of the fully kinetic Maxwell-Vlasov linear theory,
which rests upon the assumption of
an infinite homogeneous plasma and a straight uniform magnetic field,
this distribution function is parameterised  by the drift
velocity parallel to the magnetic field $u_\parallel$, the drift speed 
perpendicular to the magnetic
field $u_{\perp}$, and the thermal velocity in both directions $v_{th}$. The
thermal spread of the alpha particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

\begin{align}
f_\alpha\left(v_\parallel, v_\perp\right) &=
  f_{\alpha,\parallel}\left(v_\parallel\right)f_{\alpha,\perp}\left(v_\perp\right)\\
f_{\alpha,\parallel}\left(v_\parallel\right) &=
  \frac{\exp\left(-\frac{(v_\parallel -
  u_\parallel)^2}{v_{\parallel,th}^2}\right)}{v_{\parallel,th} \sqrt{2
  \pi}}\\
f_{\alpha,\perp}\left(v_\perp\right) &=
  \frac{1}{2\pi}\frac{\exp\left(-\frac{\left(v_\bot -
  u_\bot\right)^2}{v_{\bot,th}^2}\right)}{\left(\frac{\sqrt \pi}{2} v_{\perp,
  th} u_\bot (1 - \mathrm{erf}(-\frac{u_\bot}{v_{\perp,th}})) + \exp(-u_\bot^2 /
  v_{\perp, th}^2) \frac{v_{\perp,th}^2}{2}\right)}\\
\end{align}

Calculations are made for alpha-particles with ring-beams of two
values of pitch-angle cosine: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_{ICE}})) u_0 \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46$ m and
$\psi_0=55.2^\circ$ are the major radius and pitch angle at birth, and
$R_{ICE}=4$ m is the major radius of the emission region, see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$u_\parallel=0 u_0$ representing a ring distribution with no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here, $V_A$ is the Alfv{\'en} speed and $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is
the speed of an alpha-particle with an energy of 3.6 MeV.

The plasma parameters approximate those found in the region indicated in Fig. 3
of Ref. \cite{Cottrell1993}, which are associated with the continuous trace
signal of Fig. 2 in the same reference. The electrons and deuterons have
temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$, which corresponds a major radius of $4$ m.

Although curvature and grad-B drift terms would be present in fast
alpha-particle populations on JET, it is worth noting that they are both of
order $\rho^*$ and have been neglected in these calculations where $u_\bot \gg
\rho^* u_0$ is always the case.

Solutions to the linearised Maxwell-Vlasov system of equations are given by the
dispersion relation, Eq. \ref{eq:dettensor}, which is satisfied for sets of
$(\omega,\vec{k})$ given the distribution functions of all the species present
along the fast Alfv{\'e}n branch  on a grid of $(k_\parallel, k_\bot)$.
The hot dielectric tensor\cite{Stix}, $\vec{\vec{\epsilon}}$, Eq. \ref{eq:dielectrictensor} is calculated for each species.

@@

Integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind\cite{Stix} for both electrons and
deuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution are calculated numerically via
Gauss-Kronrod\cite{Kronrod1965} quadrature using the GaussGK.jl\cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
is used to confirm convergence of the integral values.

Since all the parallel distribution functions for all species are Maxwellians,
the parallel integrals can be calculated from the well known plasma dispersion
function\cite{Fried1961}, which is laid out in the appendix.

Bessel functions indices in Eq. \ref{eq:dielectrictensor} are summed outwards
from 0 in both the positive and negative directions
until the relative change between the current and
previous iteration is less than a tolerance of $10^{-8}$. Root finding of Eq.
\ref{eq:dettensor} in complex frequency plane is performed by an amalgamation of the
Nelder-Mead\cite{Nelder1965} optimisation method with the winding number method,
whereby if the values at the nodes of the Nelder-Mead simplex, which in this
case is just a triangle, have a winding number
of zero and therefore the simplex does not contain a root then the optimisation
procedure follows the Nelder-Mead method, otherwise the simplex bifurcates.
The child simplex that contains the root is used for the next
iteration. The optimisation iteration procedure finishes when the real extent of
the simplex is less than $10^{-4}\, \Omega_i$ and the imaginary extent is less
than $10^{-5}\,\Omega_i$, where $\Omega_i$ is the cyclotron frequency of both
the alpha-particles and deuterons since they share a charge-mass ratio.
This approach is found to be adequately robust and
faster than the Nelder-Mead method or winding number approach used in isolation
of one another. Each 512 by 512 scan in $(k_\perp, k_\parallel)$ takes between 5
and 6 minutes on a laptop.

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data
in the $(\omega,k_\parallel)$ plane where shading instead indicates growth
rates. The simplicity of the consecutive unstable cyclotron harmonics visible in
panel (a) betrays the complexity of the structure in the $(\omega,k_\parallel)$
plane. Integration over $k_\parallel$ projects the solutions onto the frequency
axis, which loses the structure and produces narrow harmonics. Figure
\ref{fig:2D_imag_zeropitch} gives $(k_\perp, k_\parallel)$ space view of the
same data as Fig. \ref{fig:2D_F12_zeropitch} as well as damped and
normal modes, except shading indicates the growth rates. Overlaid
contours indicate the real frequency in units of ion cyclotron harmonics. Only
narrow regions around the cyclotron harmonics are unstable. Overlaid straight
traces indicate propagation angle. No Doppler shift is present due to the lack
of parallel drift in the alpha-particle ring-beam. The growth rate increases
approximately linearly with cyclotron harmonic number, which is consistent with
the literature\cite{Dendy1994}. Together these figures serve to introduce a 2D
view (growth rates in $(k_\perp, k_\parallel)$ space) of traditionally 1D data
(growth rates against frequency). The importance of a non-zero $u_\parallel$ is
discussed next.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig1.pdf}
    \caption{(colour online) Unstable fast Alfv{\'e}n wave solutions to the
    linear Maxwell-Vlasov system of equations, Eq. \ref{eq:dettensor}, for
    thermal electrons, deuterons and a ring distribution of energetic
    alpha-particles i.e. when there is zero parallel drift, $u_\parallel=0$ in
    Eq. 1. (a) Upper panel: growth rate as a function of real frequency where shading
    indicates the parallel wavenumber, $k_\parallel$, in units of $\Omega_i/V_A$
    of the complex frequency solution. Consecutive cyclotron harmonics are
    increasingly unstable and are narrow due to the lack of Doppler shift 
    associated with parallel drifts. (b) Lower panel: the same solutions
    as in panel (a) in the $(\omega,k_\parallel)$ plane where shading indicates 
    growth rate. The complex structure of the growth rates that exists in
    wavenumber space is projected into sharp harmonics on the frequency axis.}
    \label{fig:2D_F12_zeropitch}
\end{figure}

@@

There are several points worth noting about Fig. \ref{fig:2D_F12_physicalpitch}.
First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Ref. \cite{Dendy1994} (also
Fig. 9 Ref. \cite{Dendy1995}) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign, which pushes the real
frequency from one side of the harmonic to the other. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation.

>>>>

\begin{abstract}
Ion cyclotron emission (ICE) originating from confined populations of fast ions
  in toroidal fusion plasmas is an important non-invasive, passive diagnostic
  for current and next generation devices. The ability to model ICE signals
  accurately is an essential step towards inferring the characteristics of
  confined energetic alpha-particles or fast neutral beam ions and background plasma.
  In this paper, the linear growth rates of the magnetoacoustic cyclotron
  instability, which is the leading explanation for ICE, are calculated in high
  resolution 2D $(k_\perp, k_\parallel)$ space for parameters corresponding to
  JET Pulse No. 26148. These calculations shed further light on the origin of
  doublet-splitting of the cyclotron harmonics as well as the perceived need to
  invoke nonlinear interactions with 1D3V particle-in-cell studies to account
  for lower harmonics with substantial amplitudes: the inclusion of signals of
  ICE from a well resolved range of propagation angles can also account for
  these effects.
\end{abstract}

@@

Observations of ion cyclotron emission (ICE) have been reported from
magnetically confined plasmas first in TFTR\cite{Cauffman1995,Dendy1995}
and JET\cite{Cottrell1988,Cottrell1993,Dendy1995}, followed by other tokamaks
PDX\cite{Heidbrink1994}, JT-60-U\cite{Ichimura2008}, DIII-D\cite{Heidbrink2011},
ASDEX\cite{DInca2014}, KSTAR\cite{Chapman2017,Chapman2018},
ASDEX-Upgrade\cite{Ochoukov2018},
TUMAN-3M\cite{Askinazi2018}, EAST\cite{Liu2020}. Stellarators\cite{Saito2013} and
FRCs\cite{Nicks2020} have also exhibited this phenomenon. Energetic ion
populations that give rise to ICE have their origins in fusion
reactions in DT plasmas\cite{Cottrell2000, Cauffman1995},
ICRH\cite{Cottrell2000}, and neutral beam injection\cite{Ochoukov2018}. ICE
arising from fusion $\alpha$-particles is a proposed route to further
the understanding of burning plasmas in ITER\cite{Dendy2015,McClements2015}.

The foremost explanation of the mechanism responsible for ICE is the
magnetoacoustic cyclotron instability (MCI)\cite{Belikov1976}, which arises from
the free energy present in energetic ions with inverted distributions in
velocity space. Theoretical understanding of ICE is based in analytical
solutions to the linearised Maxwell-Vlasov system of equations\cite{Stix} where
the energetic ion species was modelled as ring-beam for which $f(v_\parallel,
v_\perp) \propto \exp(-(v_\parallel-u_\parallel)^2/ v_{th}^2)\delta(v_\perp -
u_\bot)$ \cite{Dendy1994, McClements1996}. Here $u_\parallel$, $u_\perp$ and
$v_{th}$ are the beam velocity (the signed drift speed along the magnetic
field), ring speed (the particle speed perpendicular to the magnetic field)
and thermal velocity, respectively. Subsequent studies augmented this
understanding via nonlinear fully-kinetic\cite{Cook2013,Chapman2017} and
hybrid\cite{Carbajal2014,Reman2019} particle-in-cell (PIC) simulations in one
spatial dimension and three vector components for velocities and electromagnetic
field (i.e. 1D3V), and recently 2D3V hybrid-PIC simulations\cite{Carbajal2021}. These
computational approaches operate under similar simplifications as the analytical
treatments in that the plasma is initially spatially homogeneous and periodic
in the spatial domain.
Linearised fluid approaches that incorporate toroidal geometry and
inhomogeneous plasma and magnetic fields have shed light of the eigenmode
structure of compressional Alfv{\'e}n eigenmodes, which as with ICE are also fast
Alfv{\'e}n modes driven by energetic ions, in spherical
tokamaks\cite{Sharapov2014, Gorelenkov2016} and conventional aspect ratio
tokamaks \cite{Kolesnichenko2000, Fulop2000, Heidbrink2006}.

The effectiveness of analytical theory to reproduce the observed cyclotron
harmonic peaks in the fluctuations of the magnetic field was first explained
somewhat by 1D PIC simulations\cite{Cook2013}.
In these simulations the onset of nonlinearity
appeared to imprint the amplitudes of each of the harmonics into the signal as
they are at the end of the linear phase; each mode grows independently until 
growth is terminated for all modes when the fastest mode enters its nonlinear
phase meaning that both the amount of free energy available for instability has
decreased and that the unstable ion distribution function has been
perturbed into a more stable state. Analysis shows that only higher cyclotron
harmonics are destabilised\cite{McClements2015} to the MCI under these
conditions, however experimental
observations clearly indicate power at theoretically stable lower
harmonics\cite{Cottrell1988}. In a further step towards accurate recreation of
experimental observations, nonlinear phases of hybrid\cite{Carbajal2014} and
fully kinetic\cite{Cook2013,Chapman2017} PIC simulations went on to show that
wave-wave coupling was responsible for raising the amplitude of lower cyclotron
harmonics. Most recently, hybrid 2D simulations have begun to shed light on the
2D3V dynamics of ICE\cite{Carbajal2021}.

Finer detailed structure has been obvserved in ICE spectra, such as doublet
splitting of individual harmonics on JET\cite{Cottrell1993} and
DIII-D\cite{Thome2018} and even multiplet splitting on TUMAN-3M\cite{Askinazi2018},
ASDEX-Upgrade\cite{Ochoukov2019} and EAST\cite{Lui2020}.
This paper highlights one underlying kinetic mechinism that produces
doublet splitting due to Doppler shifted resonances and the
intricate landscape of the MCI growth rate on the fast Alfv{\'e}n dispersion
surface. In principle, splitting could be produced by grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998} or potentially any mechanism that is capable
of adding a Doppler shift to the wave-particle resonance condition. Note that
this study focusses on edge-ICE, named for the region of plasma from which the signal
arises, as opposed to core-ICE\cite{Ochoukov2018, Liu2020}.

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. In this work, I present
the growth rates in 2D wavenumber space of the MCI under the conditions of
outer mid-plane of the JET PTE\cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
LinearMaxwellVlasov.jl\cite{LinearMaxwellVlasov.jl}, written in the Julia programming
language\cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run on desktop computers
in a matter of minutes, further enabling research into this topic.

In section 2, the computational method is described for calculating
the linear growth rates of the MCI, which are presented for energetic ring-beam
alpha-particles with non-zero perpendicular thermal spread in section 3
where a comparison is ade to the observations on JET..
Results are summarised in section 4.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are found in
parallel and perpendicular wavenumber space, $(k_\bot, k_\parallel)$, for a
ring-beam distribution of energetic minority
alpha-particles and Maxwellian electrons and background thermal majority
deuteron ions. The ring-beam distribution function, Eq. 1, is a separable
function of the beam in $v_\parallel$ given by Eq. 2 and the
ring in $v_\perp$ given by Eq. 3.
In the context of the fully kinetic Maxwell-Vlasov linear theory,
which rests upon the assumption of
an infinite homogeneous plasma and a straight uniform magnetic field,
this distribution function is parameterised  by the drift
velocity parallel to the magnetic field $u_\parallel$, the drift speed 
perpendicular to the magnetic
field $u_{\perp}$, and the isotropric thermal velocity $v_{th}$. The
thermal spread of the alpha particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to the majority fuel deuterons is $1.5\times10^{-4}$.

\begin{align}
f_\alpha\left(v_\parallel, v_\perp\right) &=
  f_{\alpha,\parallel}\left(v_\parallel\right)f_{\alpha,\perp}\left(v_\perp\right)\\
f_{\alpha,\parallel}\left(v_\parallel\right) &=
  \frac{\exp\left(-\frac{(v_\parallel -
  u_\parallel)^2}{v_{\parallel,th}^2}\right)}{v_{\parallel,th} \sqrt{2
  \pi}}\\
f_{\alpha,\perp}\left(v_\perp\right) &=
  \frac{1}{2\pi}\frac{\exp\left(-\frac{\left(v_\bot -
  u_\bot\right)^2}{v_{\bot,th}^2}\right)}{\left(\frac{\sqrt \pi}{2} v_{\perp,
  th} u_\bot (1 - \mathrm{erf}(-\frac{u_\bot}{v_{\perp,th}})) + \exp(-u_\bot^2 /
  v_{\perp, th}^2) \frac{v_{\perp,th}^2}{2}\right)}
\end{align}

Calculations are made for alpha-particles with ring-beams of two
values of pitch-angle cosine: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_{ICE}})) u_0 \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46$ m and
$\psi_0=55.2^\circ$ are the major radius and pitch angle at birth, and
$R_{ICE}=4$ m is the major radius of the emission region; see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$(u_\parallel=0 u_0, u_\bot=u_0)$ representing a ring distribution with
no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here, $V_A$ is the Alfv{\'en} speed and $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is
the speed of an alpha-particle with an energy of 3.6 MeV.

The plasma parameters approximate those found at the outer mid-plane edge
$(R=4\, \mathrm{m}, Z=0\, \mathrm{m})$ where the electrons and deuterons have
temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$ at this major radius.

Although curvature and grad-B drift terms would be present in fast
alpha-particle populations on JET and these drifts would alter the wave-particle
resonance condition, it is worth noting that they are both of
order $\rho^*$ and have been neglected in these calculations where $u_\bot \gg
\rho^* u_0$ is always the case.

Solutions to the linearised Maxwell-Vlasov system of equations are given by the
dispersion relation, Eq. \ref{eq:dettensor}, which is satisfied for sets of
$(\omega,\vec{k})$ given the distribution functions of all the species present
along the fast Alfv{\'e}n branch  on a grid of $(k_\parallel, k_\bot)$.
The hot dielectric tensor\cite{Stix}, $\vec{\vec{\epsilon}}$ given by Eq.
\ref{eq:dielectrictensor} is calculated for each species, $s$,

@@

Integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind\cite{Stix} for both electrons and
deuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution given by Eq. 3 are calculated numerically via
Gauss-Kronrod\cite{Kronrod1965} quadrature using the GaussGK.jl\cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
is used to confirm convergence of the integral values.
Since all the parallel distribution functions for all species are Maxwellians,
the parallel integrals can be calculated from the well known plasma dispersion
function\cite{Fried1961}, which is laid out in the appendix.

Bessel function indices, $n$, in Eq. \ref{eq:dielectrictensor} are summed outwards
from 0 in both the positive and negative directions
until the relative change between the current and
previous iteration is less than a tolerance of $10^{-8}$. 
Root finding of Eq. \ref{eq:dettensor} in the complex frequency plane
for a given $(k_\bot, k_\parallel)$ is performed by an amalgamation of the
Nelder-Mead\cite{Nelder1965} optimisation method with the winding number method.
If the complex values at the nodes of the Nelder-Mead simplex, which in this
case is just a triangle, have a winding number
of zero and therefore the simplex does not contain a root then the optimisation
procedure follows the Nelder-Mead method, otherwise the simplex bifurcates.
The child simplex that contains the root is used for the next
iteration. The optimisation iteration procedure finishes successfully when
a simplex that encircles a root is small enough such that its extent in real
(imaginary) frequency is less than $10^{-4}\, \Omega_i$ $(10^{-5}\, \Omega_i)$
, where $\Omega_i$ is the cyclotron frequency of both
the alpha-particles and deuterons since they share a charge-mass ratio.
This approach is found to be adequately robust and
faster than either the Nelder-Mead method or winding number approach used
separately. Each 512 by 512 scan in $(k_\perp, k_\parallel)$ takes between 5
and 6 minutes on a current laptop.

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data
in the $(\omega,k_\parallel)$ plane where shading instead indicates growth
rates in units of $\Omega_i$.
The simplicity of the consecutive unstable cyclotron harmonics visible in
panel (a) betrays the complexity of the structure in the $(\omega,k_\parallel)$
plane. Integration over $k_\parallel$ projects the solutions onto the frequency
axis, which loses the structure and produces narrow harmonics. Figure
\ref{fig:2D_imag_zeropitch} gives $(k_\perp, k_\parallel)$ space view of the
same data as Fig. \ref{fig:2D_F12_zeropitch} as well as damped and
normal modes, except shading indicates the growth rates. Overlaid
contours indicate the real frequency in units of ion cyclotron harmonics. Only
narrow regions around the cyclotron harmonics are unstable. Overlaid straight
traces indicate propagation angle. No Doppler shift is present due to the lack
of parallel drift in the alpha-particle ring-beam. The growth rate increases
approximately linearly with cyclotron harmonic number, which is consistent with
the literature\cite{Dendy1994}. Together these figures serve to introduce a 2D
view (growth rates in $(k_\perp, k_\parallel)$ space) of traditionally 1D data
(growth rates against frequency). The importance of a non-zero $u_\parallel$ is
discussed next.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig1.pdf}
    \caption{(colour online) Unstable fast Alfv{\'e}n wave solutions to the
    linear Maxwell-Vlasov system of equations, Eq. \ref{eq:dettensor}, for
    thermal electrons, deuterons and a ring distribution of energetic
    alpha-particles i.e. when there is zero parallel drift, $u_\parallel=0$ in
    Eq. 1. (a) Upper panel: growth rate as a function of real frequency where shading
    indicates the parallel wavenumber, $k_\parallel$, in units of $\Omega_i/V_A$
    of the complex frequency solution. Consecutive cyclotron harmonics are
    increasingly unstable and are narrow due to the lack of Doppler shift 
    associated with parallel drifts. (b) Lower panel: the same solutions
    as in panel (a) in the $(\omega,k_\parallel)$ plane where shading indicates 
    growth rate in units of $\Omega_i$.
    The complex structure of the growth rates that exists in
    wavenumber space is projected into sharp harmonics on the frequency axis.}
    \label{fig:2D_F12_zeropitch}
\end{figure}

@@

There are several points worth noting about Fig. \ref{fig:2D_F12_physicalpitch}.
First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Ref.
\cite{Cottrell1993} for a single
signal showing multiple doublet splittings.
The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign, which pushes the real
frequency from one side of the harmonic to the other. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation.
This is hinted at by  Fig. 3 of Ref. \cite{Dendy1994} (also
Fig. 9 Ref. \cite{Dendy1995}), which shows growth rates at slices of
propagation angle that when all superimposed point towards doublet splitting,
albeit under-resolved in wavenumber space.

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions
since they arise from contributions from a range of propagation angles.
Note that Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes
more clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$k_\bot \geq 7\Omega_i/V_A$ is dominated by broader bands of instability
near to Doppler shifted cyclotron harmonics, which have
an intricate structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 13 \Omega_i/V_A$, i.e. a propagation
angle of $\sim 93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling modes and
propagation angles that are close to perpendicular.

Figure \ref{fig:2D_real_zeropitch}
shows the smoothed real frequency part of the solution for the case of
zero pitch angle cosine and is some normalised by
the high frequency expression for the fast Alfv{\'e}n wave set out in Eq.
25 of Ref. \cite{Dendy1994} (also Eq. 11 of Ref. \cite{McClements1996}),

\begin{equation}
  \omega^2 = \frac{V_A^2}{2}\left[
    k^2 + k_\parallel^2 + k^2k_\parallel^2 \frac{V_A^2}{\Omega_i^2}
  + \sqrt{
    \left(k^2 + k_\parallel^2 + k^2k_\parallel^2 \frac{V_A^2}{\Omega_i^2}\right)^2
    - 4k^2 k_\parallel^2
}
  \right],
\label{eqn:fastalfven}
\end{equation}
